Exploring mathematical objects from custom-tailored mathematical universes

### 2020-01-11 paper-filmat.tex
Beginnings of a beginning

\noindent

\section{Unsorted}

Toposes can be pictured as ``universes in which we can do mathematics in'' in
much the same way as models of set theory can be viewed in this way. In fact,
to any model~$M$ of a set theory such as~$\mathbf{ZF}$ or~$\mathbf{ZFC}$, there
is a topos~$\Set_M$ such that a statement holds in the internal language
of~$\Set_M$ if and only if it holds in~$M$.\footnote{The topos~$\Set_M$ can be
described as follows: Its objects are the elements of~$M$, that is the things
which~$M$ believes to be sets, and its morphisms are those things which~$M$
believes to be maps. The topos~$\Set_M$ validates the axioms of
$\mathbf{ETCS}$~\cite{XXX}, and for models which are not elementarily
equivalent, their associated toposes will not be equivalent.} This
includes~$V$, the trivial inner model consisting of all sets there are, in case
such a thing exists in one's chosen ontology; in this case the associated
topos~$\Set_V$ is the category of all sets, and a statement holds in~$\Set_V$
if and only if it's true in the ordinary mathematical sense.

>>>>

\noindent
Toposes can be pictured as mathematical universes in which we can do
mathematics in. Most mathematicians spend all their professional life in just a
single topos, the so-called \emph{standard topos}. However, besides the
standard topos, there is a colorful host of alternate toposes which are just as
worthy of mathematical study and in which mathematics plays out slightly
differently.

For instance, there are toposes in which the axiom of choice and the
intermediate value theorem from undergraduate calculus fail, toposes in which
any function~$\RR \to \RR$ is continuous, toposes in which infinitesimal
numbers exist and toposes whose properties depend on certain circumstances of
our physical world.

The purpose of this contribution is twofold.
\begin{enumerate}
\item We give a glimpse of the toposophic landscape, presenting several
specific toposes and exploring their peculiar properties.

\item We explicate how each topos provides a distinct lens through which the
usual mathematical objects of the standard topos can be viewed.
\end{enumerate}

Viewed through such a lens, a given mathematical object can have different
properties than it has when considered normally. In particular, it can have
better properties for the purposes of specific applications, especially if
the topos is custom-tailored to the object in question. This change of
perspective has been used in mathematical practice and demonstrates that
toposes go much beyond being logicians' testbeds. To give just a taste of what
is possible, through the lens provided by an appropriate topos, any given ring
can look like a field.

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they point the
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we could fathom the alternate
heavens of all other toposes, all embedded in a secord-order heaven.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
\end{enumerate}
This note only touches on these topics and invites further research.

\bigskip
\paragraph{Acknowledgments} We are grateful to XXX for many invaluable
discussions shaping this work and to XXX for their careful criticism of earlier
drafts. XXX conference organizer and participants

\section{Toposes as alternate mathematical universes}

Formally, a topos is a certain kind of \emph{category}, containing objects and
morphisms between those objects. For instance, the standard topos contains all
sets as its objects and all maps between sets as morphisms. 
% The term \emph{standard topos} reflects that all objects of
% mathematical study can be encoded as sets -- a group is a \emph{set} of
% elements together ...
% \begin{ex}The \emph{standard topos}~$\Set$ contains all
% sets as its objects and all maps between sets as morphisms.\end{ex}
A careful definition of the notion of a topos requires some amount
of category theory, but, as will be outlined in the following sections,
exploring the mathematical universe of a given topos does not.

Put briefly, the axioms state that a topos should share several categorical
properties with the category of sets (namely having finite limits, being
cartesian closed and having a subobject classifier); they ensure that each topos contains its
own versions of familiar mathematical objects such as natural numbers, real
numbers, groups and manifolds, and is closed under the usual constructions
such as cartesian products or quotients.

Given a topos~$\E$, we write~``$\E \models \varphi$'' to denote that a
mathematical statement~$\varphi$ \emph{holds in~$\E$}. The meaning of~``$\E \models
\varphi$'' is defined by recursion on the structure of~$\varphi$ following the
so-called \emph{Kripke--Joyal translation rules}. For instance, the rule for
translating conjunction reads
\[ \E \models (\alpha \wedge \beta) \qquad\text{iff}\qquad
  \E \models \alpha \quad\text{and}\quad \E \models \beta. \]
The remaining translation rules are more involved; we do not list them here for
the case of a general topos~$\E$, but we will state them in the next sections
for several specific toposes.

In the definition of~$\E \models \varphi$, the statement~$\varphi$ can be any
statement in the language of a general version of higher-order predicate
calculus with dependent types. In practice almost any mathematical statement
can be interpreted in a given topos.\footnote{The main exceptions are
statements from set theory, which typically make substantial use of a global
membership predicate~``$\in$''. Toposes only support a typed \emph{local}
membership predicate, where we may write~``$x \in A$'' only in the context of
some fixed sort type~$B$ such that~$x$ is of type~$B$ and~$A$ is of
type~$P(B)$, the power type of~$B$.}

It is by the Kripke--Joyal translation rules that we can access the alternate
universe of a topos. In the special case of the standard topos~$\Set$, the
definition of~``$\Set \models \varphi$'' unfolds to~$\varphi$ for any
statement~$\varphi$. Hence a statement holds in the standard topos if and only
if it holds in the usual mathematical sense.

\subsection{The logic of toposes} By their definition as special kinds of
categories, toposes are merely algebraic structures not unlike groups and
vector spaces are. Hence we need to argue why we picture toposes as
mathematical universes while we do not elevate other kinds of algebraic structures
in the same way. For us, the reason is given by the following metatheorem:

\begin{thm}Let~$\E$ be a topos and~$\varphi$ be a statement such that~$\E \models
\varphi$. If~$\varphi$ intuitionistically entails a further statement~$\psi$,
then~$\E \models \psi$.\end{thm}

XXX explain intuitionistically

XXX explain that this metatheorem allows us to \emph{reason} in a topos

XXX include something like this: A natural question is this: \emph{Which of
the familiar mathematical facts of the standard topos carry over to arbitrary
toposes~$\E$?} For any given mathematical statement~$\varphi$ and topos~$\E$,
we can unroll the definition of~$\E \models \varphi$ to try to check
whether~$\varphi$ holds in~$\E$ on an individual case-by-case basis; but it
turns out that there is a general metatheorem ...

XXX Mathematicians are familiar with the fact that the usual objects of
mathematical study are governed by the laws of ordinary \emph{classical
reasoning}. ...

\subsection{Models of set theory}

Toposes can be pictured as ``universes in which we can do mathematics in'' in
much the same way as models of set theory can be viewed in this way. In fact,
to any model~$M$ of a set theory such as~\textsc{zf} or~\textsc{zfc}, there
is a topos~$\Set_M$ such that a statement holds in the internal language
of~$\Set_M$ if and only if it holds in~$M$.\footnote{The topos~$\Set_M$ can be
described as follows: Its objects are the elements of~$M$, that is the things
which~$M$ believes to be sets, and its morphisms are those things which~$M$
believes to be maps. The topos~$\Set_M$ validates the axioms of
$\mathbf{ETCS}$~\cite{XXX}, and for models which are not elementarily
equivalent, their associated toposes will not be equivalent.} This
includes~$V$, the trivial inner model consisting of all sets there are, in case
such a thing exists in one's chosen ontology; in this case the associated
topos~$\Set_V$ is the category of all sets, and a statement holds in~$\Set_V$
if and only if it's true in the ordinary mathematical sense.
